\vspace{-1.5cm}\huge Words That Matter:\\ Vocabulary Pruning in ModernBERT

### 2025-06-04 paper/main.tex
dbscan results clustering

\section{Introduction}
Language models need to be efficient for optimal deployment.
Efficiency can be assessed in several dimensions, including inference latency, throughput, energy consumption, parameter count, and memory usage.
In this work, we study how to optimally reduce the memory footprint of language models by removing unimportant vocabulary tokens from the embedding layer, resulting in faster model loading, reduced deployment costs, and improved accessibility for edge devices or resource-constrained environments.

@@

ModernBERT stands out as one of the most lightweight and optimized encoder-only transformers available. It integrates several architectural and training innovations while maintaining robust downstream performance. Despite these advancements, we investigate whether ModernBERT's efficiency can be further enhanced through parameter pruning. Given ModernBERT's inherent efficiency, optimal pruning strategies could yield exceptionally efficient models.

Specifically, we explore task-specific weight pruning, as ModernBERT is intended to be finetuned on specific tasks for optimal behavior (unlike more general-purpose decoder-only models that can utilize in-context learning). We hypothesize that simple task-specific pruning of the embedding matrix can be highly efficient, as many tokens can be irrelevant for any specific downstream task, and the embedding layers counts for a significant portion ($\sim$25\%) of the total parameters.

We explore a series of simple vocabulary pruning heuristics on a range of Natural Language Processing tasks, including linguistic acceptability, sentiment analysis, paraphrase detection, semantic similarity, question answering, and natural language inference, as covered by the GLUE benchmark \cite{wang2018glue}. For any task, we first apply vocabulary pruning to the base ModernBERT model and subsequently finetune the pruned model. We compare these vocabulary pruning techniques with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers \cite{li2023losparse}.

@@

Our results indicate that the memory footprint of light-weight models, such as ModernBERT, can still be significantly improved.  We open-source all our code and pruned models.

\section{Related work}
Model compression techniques enhance the efficiency of large language models (LLMs) by reducing their size and computational requirements.
% 
These techniques can be broadly categorized into four primary methods~\cite{wan2023efficient}:
(i) \textbf{quantization}, which reduces the numerical precision of model weights;
(ii) \textbf{parameter pruning}, which eliminates redundant or less important connections and neurons to create sparser models with fewer parameters;
(iii) \textbf{low-rank approximation}, which decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters~\cite{Hu2021LoRA}; and
(iv) \textbf{knowledge distillation} transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size ~\cite{Hinton2015Distillation}.

%can you give an example here?
%Misschien niet nodig voor de paper maar heel goed om je expertise te tonen in de full thesis
The model compression techniques described above are complementary and can be combined to achieve optimal compression.
This work focuses on parameter pruning techniques to reduce model size while preserving performance. The rest of this section gives an overview of the most important parameter pruning techniques.

@@

These encoder-layer pruning methods differ in their granularity and strategy—targeting individual weights, attention heads, or full layers. LoSparse stands out as a strong baseline for encoder-level pruning~\cite{pmlr-v235-li24bi}~\cite{huang2023awesome}~\cite{ren2025llmcompression}.

@@

To our knowledge, this is the first study to (i) rigorously evaluate a range of importance heuristics in a single framework, (ii) incorporate OOV‑aware mapping to preserve semantic fidelity of pruned tokens, and (iii) apply vocabulary pruning atop ModernBERT—a state‑of‑the‑art lightweight encoder—thereby demonstrating the practical benefits of embedding‑layer reduction in resource‑constrained settings.

@@

\textbf{Clustering-Based}
preserves semantic diversity across the vocabulary by grouping similar tokens together and keeping only one representative from each group. It leverages the property that embeddings pointing in similar directions (high cosine similarity) typically have similar meanings\cite{mikolov2013efficient}. 
This process ensures maximal semantic coverage with a reduced vocabulary size.

@@

%For example, the important tokens identified in the SST-2 dataset differ markedly between methods. Frequency-based approaches naturally prioritize common words like articles and prepositions, while attention-based analysis highlights sentiment-laden terms that are semantically relevant to the classification task.
% \begin{figure*}[t]
% \centering
% \begin{minipage}{0.48\textwidth}
%   \centering
%   \includegraphics[width=\textwidth]{images/wordcloud_frequency.png}
% \end{minipage}%
% \hfill
% \begin{minipage}{0.48\textwidth}
%   \centering
%   \includegraphics[width=\textwidth]{images/wordcloud_attention.png}
% \end{minipage}
% \caption{Comparison of top-ranked tokens for SST-2 task using frequency-based (left) versus attention-based (right) importance metrics.}
% \label{fig:wordcloud-comparison}
% \end{figure*}
% The visualization in Figure \ref{fig:wordcloud-comparison} demonstrates how attention-based pruning better captures task-relevant information, preserving tokens that carry greater semantic weight for the specific downstream task.

\textbf{Attention-Based}
leverages the attention mechanism of transformer models to identify potentially important tokens. The underlying principle is that tokens receiving higher attention during inference might be those the model relies on more when making predictions, offering a model-based perspective on token importance\cite{guo2024attention}. While attention scores do not necessarily imply causal relevance, they provide a task-specific signal that differs from simple statistical measures. The approach first finetunes the base model on the target task to learn task-specific attention patterns, then processes the dataset through this model to capture attention matrices from all layers and heads. It aggregates the attention each token receives across all its occurrences, and finally normalizes scores by token frequency to avoid bias toward common tokens.

@@

% The clustering-based Out-Of-Vocabulary (OOV) handling process extracts embeddings for all pruned tokens and applies K-means clustering to group semantically similar tokens together. From each cluster, the token closest to the centroid is selected as a representative. During inference, when an OOV token is encountered, it is mapped to its assigned representative.
The clustering-based Out-Of-Vocabulary (OOV) handling process extracts embeddings for all pruned tokens and applies K-means clustering to group semantically similar tokens together. From each cluster, the token closest to the centroid is selected as a representative. While alternative clustering methods (e.g., DBSCAN \cite{wang2019dbscan}) and representative selection strategies (e.g., averaging embeddings) could be explored, we focus on this approach for its simplicity and effectiveness. During inference, when an OOV token is encountered, it is mapped to its assigned representative.
The number of clusters (k) offers a tunable parameter to balance compression rate against semantic precision. While this approach adds a few more tokens back into the vocabulary, the intuition is that it can recover most of the performance lost during pruning.

@@

\textbf{As baselines,} 
we establish three baselines to evaluate our vocabulary pruning techniques: 
(i) \textit{ModernBERT (Full)} serves as the upper performance bound, using the complete 50,368-token vocabulary and all encoder parameters, finetuned on the training dataset of each task; 
(ii) \textit{LoSparse} provides a complementary encoder-layer pruning baseline that uses low-rank and sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer; 
(iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes tokens not observed in finetuning data, achieving variable parameter reduction (4.79-22.61\%) depending on the dataset's vocabulary coverage.

@@

\section{Future Work and Limitations}
Several limitations constrain our approach. 
Our clustering-based methods for both token importance estimation and OOV handling rely on K-means clustering, while alternative methods such as DBSCAN~\cite{wang2019dbscan} might better capture complex semantic relationships in embedding spaces without requiring pre-specified cluster numbers.
Similarly, our attention-based token importance method relies solely on attention scores, but recent work suggests this may be incomplete, as investigation into value vector norms reveals notably non-uniform patterns~\cite{guo2024attention}. Future work should explore Value-Aware Token Pruning (VATP) approaches that combine attention scores with the $\ell_1$ norm of value vectors for more comprehensive token importance evaluation.

>>>>

\section{Introduction}
Language models need to be efficient for optimal deployment.
Efficiency can be assessed in several dimensions, including inference latency, throughput, energy consumption, parameter count, and memory usage\cite{yuan2025efficientllm}.
In this work, we study how to optimally reduce the memory footprint of language models by removing unimportant vocabulary tokens from the embedding layer, resulting in faster model loading, reduced deployment costs, and improved accessibility for edge devices or resource-constrained environments.

@@

ModernBERT \cite{warner2024modernbert} stands out as one of the most lightweight and optimized encoder-only transformers available. It integrates several architectural and training innovations while maintaining robust downstream performance. Despite these advancements, we investigate whether ModernBERT's efficiency can be further enhanced through parameter pruning. Given ModernBERT's inherent efficiency, optimal pruning strategies could yield exceptionally efficient models.

Specifically, we explore task-specific weight pruning, as ModernBERT is intended to be finetuned on specific tasks for optimal behavior (unlike more general-purpose decoder-only models that can utilize in-context learning). We hypothesize that simple task-specific pruning of the embedding matrix can be highly efficient, as many tokens can be irrelevant for any specific downstream task, and the embedding layers counts for a significant portion ($\sim$25\%) of the total parameters \cite{yang2022taskspecific}.

We explore a series of simple vocabulary pruning heuristics on a range of Natural Language Processing tasks, including linguistic acceptability, sentiment analysis, paraphrase detection, semantic similarity, question answering, and natural language inference, as covered by the GLUE benchmark \cite{wang2018glue}. For any task, we first apply vocabulary pruning to the base ModernBERT model and subsequently finetune the pruned model. We compare these vocabulary pruning techniques with LoSparse \cite{li2023losparse}, a state-of-the-art gradient-based pruning technique that targets the encoder layers.

@@

Our results indicate that the memory footprint of light-weight models, such as ModernBERT, can still be significantly improved.  We open-source all our code.

\section{Related work}
Model compression techniques enhance the efficiency of large language models (LLMs) by reducing their size and computational requirements.
% 
These techniques can be broadly categorized into four primary methods~\cite{wan2023efficient, park2024comprehensive, cheng2017survey, mishra2020survey, xailient2021compression}:
(i) \textbf{quantization}, which reduces the numerical precision of model weights \cite{gholami2021survey};
(ii) \textbf{parameter pruning}, which eliminates redundant or less important connections and neurons to create sparser models with fewer parameters \cite{cheng2023survey};
(iii) \textbf{low-rank approximation}, which decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters~\cite{Hu2021LoRA}; and
(iv) \textbf{knowledge distillation} transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size ~\cite{Hinton2015Distillation}.

%can you give an example here?
%Misschien niet nodig voor de paper maar heel goed om je expertise te tonen in de full thesis
The model compression techniques described above are complementary and can be combined to achieve optimal compression \cite{muralidharan2024compact, han2015deep, choukroun2019low}.
This work focuses on parameter pruning techniques to reduce model size while preserving performance. The rest of this section gives an overview of the most important parameter pruning techniques.

@@

These encoder-layer pruning methods differ in their granularity and strategy—targeting individual weights, attention heads, or full layers. LoSparse stands out as a strong baseline for encoder-level pruning~\cite{pmlr-v235-li24bi, huang2023awesome, ren2025llmcompression}.

@@

To our knowledge, this is the first study to (i) rigorously evaluate a range of importance heuristics in a single framework, (ii) incorporate OOV‑aware mapping to preserve semantic fidelity of pruned tokens, and (iii) apply vocabulary pruning atop ModernBERT thereby demonstrating the practical benefits of embedding‑layer reduction in resource‑constrained settings.

@@

\textbf{Clustering-Based}
preserves semantic diversity across the vocabulary using DBSCAN~\cite{wang2019dbscan} clustering to group similar token embeddings and keeping one representative per cluster. This leverages the property that embeddings with high cosine similarity typically have similar meanings\cite{mikolov2013efficient}, ensuring maximal semantic coverage with reduced vocabulary size.

@@

\textbf{Attention-Based}
leverages the attention mechanism of transformer models to identify potentially important tokens. The underlying principle is that tokens receiving higher attention during inference might be those the model relies on more when making predictions, offering a model-based perspective on token importance\cite{guo2024attention, kim2022learned, goyal2020powerbert, chen2021earlybert}. While attention scores do not necessarily imply causal relevance, they provide a task-specific signal that differs from simple statistical measures. The approach first finetunes the base model on the target task to learn task-specific attention patterns, then processes the dataset through this model to capture attention matrices from all layers and heads. It aggregates the attention each token receives across all its occurrences, and finally normalizes scores by token frequency to avoid bias toward common tokens.

@@

% The clustering-based Out-Of-Vocabulary (OOV) handling process extracts embeddings for all pruned tokens and applies K-means clustering to group semantically similar tokens together. From each cluster, the token closest to the centroid is selected as a representative. During inference, when an OOV token is encountered, it is mapped to its assigned representative.
The clustering-based Out-Of-Vocabulary (OOV) handling process extracts embeddings for all pruned tokens and applies K-means clustering to group semantically similar tokens together. From each cluster, the token closest to the centroid is selected as a representative. While alternative clustering methods (e.g., DBSCAN) and representative selection strategies (e.g., averaging embeddings) could be explored, we focus on this approach for its simplicity and effectiveness. During inference, when an OOV token is encountered, it is mapped to its assigned representative.
The number of clusters (k) offers a tunable parameter to balance compression rate against semantic precision. While this approach adds a few more tokens back into the vocabulary, the intuition is that it can recover most of the performance lost during pruning.

@@

\textbf{As baselines,} 
we establish three baselines to evaluate our vocabulary pruning techniques: 
(i) \textit{ModernBERT (Full)} serves as the upper performance bound, using the complete 50,368-token vocabulary and all encoder parameters, finetuned on the training dataset of each task; 
(ii) \textit{LoSparse} provides an encoder-layer pruning baseline that uses low-rank and sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer; 
(iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes tokens not observed in finetuning data, achieving variable parameter reduction (4.79-22.61\%) depending on the dataset's vocabulary coverage.

@@

\section{Future Work and Limitations}
While our methods show promise, several limitations remain.
Our clustering-based method for OOV handling relies on K-means clustering, while alternative methods such as DBSCAN~\cite{wang2019dbscan} might better capture complex semantic relationships in embedding spaces without requiring pre-specified cluster numbers.
Similarly, our attention-based token importance method relies solely on attention scores, but recent work suggests this may be incomplete, as investigation into value vector norms reveals notably non-uniform patterns~\cite{guo2024attention}. Future work should explore Value-Aware Token Pruning (VATP) approaches that combine attention scores with the $\ell_1$ norm of value vectors for more comprehensive token importance evaluation.

### 2025-06-04 paper/tables/results.tex
dbscan results clustering

\begin{table*}[h]
\centering
\scriptsize
\setlength{\tabcolsep}{9pt}
\begin{tabular}{l@{\hspace{25pt}}ccccccccc}
\toprule
& \multicolumn{2}{c}{\textbf{Single Sentence}} & \multicolumn{3}{c}{\textbf{Paraphrase and Similarity}} & \multicolumn{3}{c}{\textbf{Natural Language Inference}} & \multirow{2}{*}{\makebox[-20pt][c]{\vrule width 0.5pt height 175pt}\hspace{35pt}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-6} \cmidrule(lr){7-9}
\textbf{Method} & \textbf{SST-2} & \textbf{CoLA} & \textbf{MRPC} & \textbf{STS-B} & \textbf{QQP} & \textbf{MNLI} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
\midrule
\multicolumn{10}{l}{\textbf{Baseline}} \\
ModernBERT (Full) & 0.951 & 0.632 & 0.89 & 0.917 & 0.917 & 0.881 & 0.939 & 0.643 & 0.846 \\
\midrule
\multicolumn{10}{l}{\textbf{Vocabulary Pruning}} \\
Train Tokens Only & 0.950 & 0.630 & 0.861 & 0.917 & 0.917 & 0.883 & 0.915 & 0.639 & 0.839 \\
\textit{Parameter Reduction} & \textit{18.85\%} & \textit{22.61\%} & \textit{18.56\%} & \textit{18.76\%} & \textit{4.79\%} & \textit{6.74\%} & \textit{6.42\%} & \textit{17.06\%} & \textit{14.22\%} \\
\\ [-6pt]
\hdashline
\\[-6pt]
Random Selection & 0.911 & 0.470 & 0.798 & 0.845 & 0.780 & 0.504 & 0.669 & 0.566 & 0.693 \\
Clustering-Based & 0.899 & 0.051 & 0.714 & 0.786 & 0.774 & 0.501 & 0.510 & 0.566 & 0.600 \\
Attention-Based & \textbf{0.947} & 0.589 & \textbf{0.864} & 0.885 & 0.763 & 0.683 & 0.791 & 0.578 & 0.763 \\
Simple Frequency-Based & 0.943 & 0.467 & 0.812 & 0.905 & 0.904 & 0.858 & 0.902 & 0.546 & 0.792 \\
TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & 0.909 & 0.610 & 0.805 \\
\\ [-6pt]
\hdashline
\\[-6pt]
Simple Frequency + OOV & 0.938 & 0.540 & 0.828 & \textbf{0.906} & \textbf{0.915} & 0.858 & 0.907 & 0.615 & 0.813 \\
TF-IDF + OOV & 0.934 & \textbf{0.615} & 0.834 & 0.903 & 0.912 & 0.858 & \textbf{0.910} & \textbf{0.635} & \textbf{0.825} \\
\textit{Parameter Reduction} & \textit{20.25\%} & \textit{23.27\%} & \textit{20.02\%} & \textit{20.18\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.48\%} \\
\midrule
\multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
LoSparse & 0.935 & 0.525 & 0.856 & 0.887 & 0.915 & 0.871 & 0.906 & 0.614 & 0.814 \\
\textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} \\
\bottomrule
\end{tabular}
\caption{Performance on GLUE dev set. ModernBERT is fine-tuned separately for each task. Scores are accuracies except for CoLA (Matthew's correlation), and STS-B (Pearson correlation). Notation "+ OOV" indicates pruning with out-of-vocabulary clustering. At 20\% parameter reduction, TF-IDF + OOV maintains 97.6\% of original performance and outperforms LoSparse (0.824 vs 0.810 average score). OOV handling improves results by 1.9 percentage points on average.}

>>>>

\begin{table*}[h]
\centering
\scriptsize
\setlength{\tabcolsep}{9pt}
\begin{tabular}{l@{\hspace{25pt}}ccccccccc}
\toprule
& \multicolumn{2}{c}{\textbf{Single Sentence}} & \multicolumn{3}{c}{\textbf{Paraphrase and Similarity}} & \multicolumn{3}{c}{\textbf{Natural Language Inference}} & \multirow{2}{*}{\makebox[-20pt][c]{\vrule width 0.5pt height 175pt}\hspace{35pt}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-6} \cmidrule(lr){7-9}
\textbf{Method} & \textbf{SST-2} & \textbf{CoLA} & \textbf{MRPC} & \textbf{STS-B} & \textbf{QQP} & \textbf{MNLI} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
\midrule
\multicolumn{10}{l}{\textbf{Baseline}} \\
ModernBERT (Full) & 0.951 & 0.632 & 0.89 & 0.917 & 0.917 & 0.881 & 0.939 & 0.643 & 0.846 \\
\midrule
\multicolumn{10}{l}{\textbf{Vocabulary Pruning}} \\
Train Tokens Only & 0.950 & 0.630 & 0.861 & 0.917 & 0.917 & 0.883 & 0.915 & 0.639 & 0.839 \\
\textit{Parameter Reduction} & \textit{18.85\%} & \textit{22.61\%} & \textit{18.56\%} & \textit{18.76\%} & \textit{4.79\%} & \textit{6.74\%} & \textit{6.42\%} & \textit{17.06\%} & \textit{14.22\%} \\
\\ [-6pt]
\hdashline
\\[-6pt]
Random Selection & 0.911 & 0.470 & 0.798 & 0.845 & 0.780 & 0.504 & 0.669 & 0.566 & 0.693 \\
Clustering-Based & 0.894 & 0.188 & 0.752 & 0.696 & 0.871 & 0.683 & 0.833 & 0.566 & 0.685 \\
Attention-Based & \textbf{0.947} & 0.589 & \textbf{0.864} & 0.885 & 0.763 & 0.683 & 0.791 & 0.578 & 0.763 \\
Simple Frequency-Based & 0.943 & 0.467 & 0.812 & 0.905 & 0.904 & 0.858 & 0.902 & 0.546 & 0.792 \\
TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & 0.909 & 0.610 & 0.805 \\
\\ [-6pt]
\hdashline
\\[-6pt]
Simple Frequency + OOV & 0.938 & 0.540 & 0.828 & \textbf{0.906} & \textbf{0.915} & 0.858 & 0.907 & 0.615 & 0.813 \\
TF-IDF + OOV & 0.934 & \textbf{0.615} & 0.834 & 0.903 & 0.912 & 0.858 & \textbf{0.910} & \textbf{0.635} & \textbf{0.825} \\
\textit{Parameter Reduction} & \textit{20.25\%} & \textit{23.27\%} & \textit{20.02\%} & \textit{20.18\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.48\%} \\
\midrule
\multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
LoSparse & 0.935 & 0.525 & 0.856 & 0.887 & 0.915 & 0.871 & 0.906 & 0.614 & 0.814 \\
\textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} \\
\bottomrule
\end{tabular}
\caption{Performance on GLUE dev set. ModernBERT is fine-tuned separately for each task. Scores are accuracies except for CoLA (Matthew's correlation), and STS-B (Pearson correlation). Notation "+ OOV" indicates pruning with out-of-vocabulary clustering. At 20\% parameter reduction, TF-IDF + OOV maintains 97.6\% of the original performance and outperforms LoSparse (0.825 vs 0.810 average score). OOV handling improves results by 2.0 percentage points on average.}
